The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-08-17 manuscript/rebuttal.Rmd

> 2. Expanding on Current Shortcomings: The article mentions that current practices are not meeting the
> goal of reproducibility. An expanded discussion on the ramifications of these shortcomings on
> scientific progress could accentuate the urgency and significance of addressing this issue.

In the introduction we have spoken to the consequences of poor computational research, and emphasize
the urgency of this issue (which complements the case study summary above):

@@

> The ramifications of irreproducible and unreliable research include misleading the community,
> wasting research funds, slowing scientific progress, eroding public confidence in science and
> tarnishing the reputation of associated institutions and colleagues.
> In certain cases, irreproducible bioinformatics has the potential to place patient safety at risk.

> 3. Focus on Implementation Challenges: While the article acknowledges the importance of certain
> practices, an increased focus on potential obstacles researchers may face when trying to implement
> these suggestions, as well as strategies to overcome these challenges, would be insightful.

We have re-written the Challenges section to highlight some of the barriers slowing the
adoption of best practices, with reference to previous contributions on the topic.
These are structured around what different stakeholders can do to improve the situation (including
funders, journals, institutions and individual researchers).

> 4. Clear Structure: Consider organizing the article into clearly defined sections such as
> 'Introduction', 'Background', 'Recommendations', 'Challenges', and 'Conclusion'.
> This could significantly enhance readability and comprehensibility.

The subheading structure has been amended accordingly.

@@

* Key points

> 5. Engaging Conclusion: Reworking the conclusion to be more impactful could be beneficial. Instead of
> merely summarizing the recommendations, it would be useful to indicate the potential trajectory of
> computational research if these recommendations are adopted, as well as highlighting the potential
> risks if they aren't.

Response: We have completely rewritten the conclusion.
Instead of reiterating the five pillars, we focus on the motivation for adoption:

> Large scale datasets are becoming commonplace in medicine and life sciences thanks to the commoditisation
> of platform technologies like DNA sequencing, mass spectrometry and advanced imaging systems.
> These big datasets are enabling breakthrough discoveries, but due to their size and complexity, require
> significant expertise to analyze and distill key observations.
> The measures proposed here are designed to enhance transparency and foster the reuse of data and code;
> reducing research waste.
> Transparency makes it easier for problems to be identified, so that erroneous findings are less likely to
> pollute the scientific literature.
> Researchers, their institutions, publishers and funders each have a major role to play in fighting the
> reproducibility crisis by encouraging highly reproducible research practices.

### 2023-08-17 manuscript/5pillars.Rmd
reviewer 2 comments addressed I think

Containers are an attempt to resolve some of the downsides of VMs.
Container images can be thought of as similar to VMs, but are more lightweight as they
share parts of the host operating system [@Nust2020-cr].
In the example shown in **Figure 4**, the five containerised applications share the same operating
system, while the three VM applications each involve their own operating system which incurs a
significant performance overhead @noauthor_2021-lt.
Therefore, running workflows in containers incurs only a small reduction in performance as compared
to running directly on the host system @Di_Tommaso2015-pw.
Container images are highly portable because containers include everything needed to run the
application, including the system tools and libraries, ensuring that the environment is consistent
across different systems.
For example, the most popular containerisation system, Docker [@Merkel2014-ek], makes it
possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
promise of reproducibility.
While there are several alternatives such as Podman, Rkt or Apptainer/Singularity, Docker remains the most widely used
containerisation system for web development and research.
Docker now has a large community of users, extensive documentation and
a vast collection of pre-built container images in the DockerHub registry.
Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
and typically within a few minutes.
This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
reproducibility [@Mangul2019-ei;@Mangul2019-xp;@Krafczyk2021-xv].

@@

Although the technological tools underlying these best practices have existed for a decade,
progress on implementing them into computational research publications has been minimal.

Researchers face several hurdles in adopting these.

* The "publish-or-perish" incentive structures in academia reward prolific rather than scrupulous
authors.
Funding bodies, institutions and journals generally prefer to focus on novel research directions rather
than meticulous confirmation of previous findings [@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id;
@Munafo2017-ty; @Diong2021-is].

* A lack of systematic reproducibility/replication studies in bioinformatics remains small
[@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], so the severity of the reproducibility
problem isn't well appreciated.

* Journal standards are slow to progress.
Our observations indicate that sharing of data and code remains optional for most medical
journals in PubMed, but in the field of computational biology, providing these resources is
becoming standard practice.

* A lack of documentated protocols and tutorial materials for highly-reproducible analysis makes it
difficult for researchers to confidently update their code development practices.

* An overall lack of coding/data science expertise among life science researchers and new graduates.
Due to strong demand from industry and government, academia has a deficit of workers fluent in data
science [@Committee_on_Envisioning_the_Data_Science_Discipline_The_Undergraduate_Perspective2018-pe].
This makes it challenging for organisations to recruit capable data analysts and cultivate a
community of practice.

Stodden et al have outlined some practical recommendations for tackling these problems [@Stodden2016-aj].

Overcoming these obstacles will require a multi-pronged strategy and effort from different stakeholders.

* What researchers can do.

* What journals can do.

* What insitutions can do.

* What universities can do.

* What funders can do.

* Incentive structures will need to change.
How we measure and value research achievement.
A change in culture or research assessment.
How we fund research data analytics and infrastructure.
Value rigour as well as prolificness.

* Journals need to progress towards publishing reproducible workflows.
Peng 2011 speaks about the lack of infrastructure for disseminating reproducible research.

* Enhanced training.
A lack of technical computer skills in many research groups is another culprit.
Specific workshop-based training in these areas could help [@Williams2023-ic].
The [Software Carpentry](https://software-carpentry.org/) initiative is focused on this mission
[@Wilson2014-mh].
Undergraduate and postgraduate courses in the life sciences would benefit from enhanced training in data
science.

* Enhanced documentation, protocols and tutorials.
The most scalable solution would be to provide comprehensive beginner-friendly step-by-step
guides for frequently used routines.
For example, our group has just released such a protocol that walks novice users through the process
of implementing functional enrichment analysis of gene expression data with the five pillars principles
@Ziemann2023-nm.
More such protocols are needed to make extreme reproducibility mainstream.
A compendium of publicly available learning materials around the five pillar concepts are
provided in the supplementary information.

* What individuals can do.

* Overall change in culture.
Recognition that life sciences is a data intensive discipline.

@@

Large scale datasets are becoming commonplace in medicine and life sciences thanks to the commoditisation
of platform technologies like DNA sequencing, mass spectrometry and advanced imaging systems.
These big datasets are enabling breakthrough discoveries, but due to their size and complexity, require
significant expertise to analyze and distill key observations.
The measures proposed here are designed to enhance transparency and foster the reuse of data and code.
Such reuse reduces waste and enhances the cost-effectiveness of research.
The transparency provided by sharing data, code and documentation makes it easier for problems to be
identified, so that erroneous findings are less likely to pollute the scientific literature.
Researchers, their institutions, publishers and funders each have a major role to play in fighting the
reproducibility crisis by encouraging highly reproducible research practices.

https://bmcresnotes.biomedcentral.com/articles/10.1186/s13104-021-05875-3#Sec2

>>>>

Containers are an attempt to resolve some of the downsides of VMs.
Container images can be thought of as similar to VMs, but are more lightweight as they
share parts of the host operating system [@Nust2020-cr].
In the example shown in **Figure 4**, the five containerised applications share the same operating
system, while the three VM applications each involve their own operating system which incurs a
significant performance overhead @noauthor_2021-lt.
Therefore, running workflows in containers incurs only a small reduction in performance as compared
to running directly on the host system @Di_Tommaso2015-pw.
Container images are highly portable because containers include everything needed to run the
application, including the system tools and libraries, ensuring that the environment is consistent
across different systems.
For example, the most popular containerisation system, Docker [@Merkel2014-ek], makes it
possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
promise of reproducibility.
While there are several alternatives such as Podman, Rkt or Apptainer/Singularity, Docker remains the most widely used
containerisation system for web development and research.
Docker now has a large community of users, extensive documentation and
a vast collection of pre-built container images in the DockerHub registry.
Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
and typically within a few minutes.
This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
reproducibility [@Mangul2019-ei; @Mangul2019-xp; @Krafczyk2021-xv].

@@

Although the technological tools underlying these best practices have existed for a decade,
progress on implementing them into computational research publications has been minimal.
There are very many challenges, some of which have been discussed before
[ @Peng2011-ia; @Stodden2016-aj; @Munafo2017-gf ; @Diaba-Nuhoho2021-nt ].

Due to the relative lack of systematic reproducibility/replication studies in bioinformatics,
the severity of the reproducibility problem isn't well appreciated.
The studies that have been conducted point to reproducibility rates of ~10% in computational
life sciences [@Ioannidis2009-hv;@Zaringhalam2020-ms; @Samuel2022-ha], but further systematic studies
are required to bring attention to the issue.

Journals are partly responsible, as their editorial policies influence author behaviours.
Except for a few outliers, verification of computational reproducibility is not in the scope of peer
review, but this should be considered.
As others have noted, it is in the interests of funders to provide the infrastructure to facilitate this
[@Peng2011-ia].

Another hindrance to reproducibility more broadly is the fixation on novel research directions rather
than meticulous confirmation of previous findings [@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id;
@Munafo2017-ty; @Diong2021-is].
This is a direct result of research assessment policies that incentivize journal-level prestige
metrics and number of publications rather than the rigour [@Munafo2017-gf].
A change in research assessment ("eg: track-record") that recognises and credits contributions toward
reproducibility and rigor would help decrease funds wasted on sloppy research.

Individual researchers also face difficulties.
Life science graduates often receive inadequate training in data science that places them in a
disadvantage considering the data intensive nature of contemporary biology.
This is compounded by the strong demand for data science experts in industry and government which
leads to a deficit of academics fluent in data science
[@Committee_on_Envisioning_the_Data_Science_Discipline_The_Undergraduate_Perspective2018-pe].
This makes it challenging for organisations to recruit and retain capable data analysts and cultivate
a community of practice.
Addressing this problem is not straightforward.
It requires ehancement of data science skills in undergraduate life science courses, investment in
upskilling the current workforce and offering working conditions that are competitive with other sectors.
Specific workshop-based training for current researchers could help [@Williams2023-ic].
The [Software Carpentry](https://software-carpentry.org/) initiative is focused on this mission
[@Wilson2014-mh].

A lack of documentated protocols and tutorial materials for highly reproducible analysis makes it
difficult for researchers to confidently update their code development practices.
Expert bioinformaticians can help by publishing comprehensive beginner-friendly step-by-step
guides for frequently used routines.
This approach is highly scalable and the reach of open-access publishing ensures these materials are
broadly accesible.
For example, our group has just released such a protocol that walks novice users through the process
of implementing functional enrichment analysis of gene expression data with the five pillars principles
@Ziemann2023-nm.
More guides like this are needed to make extreme reproducibility mainstream.
A compendium of publicly available learning materials around the five pillar concepts are
provided in the supplementary information.

@@

Large scale datasets are becoming commonplace in medicine and life sciences thanks to the commoditisation
of platform technologies like DNA sequencing, mass spectrometry and advanced imaging systems.
These big datasets are enabling breakthrough discoveries, but due to their size and complexity, require
significant expertise to analyze and distill key observations.
The measures proposed here are designed to enhance transparency and foster the reuse of data and code;
reducing research waste.
Transparency makes it easier for problems to be identified, so that erroneous findings are less likely to
pollute the scientific literature.
Researchers, their institutions, publishers and funders each have a major role to play in fighting the
reproducibility crisis by encouraging highly reproducible research practices.
